Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-05-06 paper1_jiturra.Rmd
update Alejo

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | The escalating economic inequality and the current sanitary crisis have threatened social solidarity among citizens. However, these consequences have been unevenly experienced across the social structure where the upper and intermediate classes, benefiting from privileged access to social resources, have shown resilience. Conversely, working-class families have faced deteriorating material conditions, leading to a heightened sense of marginalization and increasing the demand for welfare support. Particularly, social resources through network ties have been suggested to have a direct link with an individual's welfare through instrumental and expressive outcomes, such as providing information or help in moments of need. Consequently, it has been argued that social isolation not only plays a role in terms of social resources, but higher levels of segregation can bolster opinions, suggesting that being isolated from other social classes can polarise attitudes mainly in the working and upper-middle classes. In this regard, based on the ISSP 2017 -  Social Networks (N=31,191),  multilevel estimations suggest that the influence of network homogeneity is conditional to social class, where the working and intermediate classes with high homogeneous networks hold higher redistributive preferences than the upper classes. Additionally, income inequality has been shown to decrease the polarizing effect of network segregation, mainly among the upper classes, when overall inequality is high. In sum, this study highlights the role of both social class and network segregation together to understand redistributive preferences, as well as their implications for policies and cohesion.
  | 
  | **Keywords**: network segregation, redistributive preferences, economic inequality
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

Rising economic inequality over the past decades [@piketty_capital_2014] and the recent sanitary crisis have raised citizens' concerns about economic inequality, pushing governments to enact redistributive policies, which has tested the capacity of welfare systems to provide sufficient risk protection for their citizens [@breznau_welfare_2021]. However, the privileged access to resources held by the upper and intermediate classes has consolidated their resilience capacity in times of greater economic hardship, allowing them to maintain a relatively stable standard of living, in contrast to the decline in the material conditions of working-class families [@moawad_myth_2023]. In this regard, it has been claimed that economic gaps in economic resources also have a greater effect on social relations. Studies on the role of economic inequality have argued that social bonding can be threatened when economic inequality increases, undermines how people perceive each other, their opportunities to participate in social life, and the capacity to knit networks that contribute to social integration [@lancee_income_2012;@letki_getting_2015]. Therefore, experiencing segregated networks is one of the consequences of greater economic gaps as it represents greater distance between the different social classes  [@pichler_patterns_2007;@otero_space_2022]. In this regard, it has been argued that segregation can threaten social cohesion and solidarity [@joye_measuring_2019], which in turn can undermine public support for welfare schemes and redistribution [@baldassarri_diversity_2020].

The sociological tradition has understood social classes through the labor market situation of occupations in terms of employment relations, skill level, and authority within the workplace [@Erikson1992;@Erikson1979]. Likewise, classes represent relations in the labor market and trace lines in terms of status relations as the foundations for opportunities for action and the consolidation of lifestyles [@weber_economy_1978]. 
In this matter, the class situation of individuals and households also trace distinctions in various domains of social life, such as cultural consumption behavior [@Chan2007;@torche_social_2007], ideological preferences [@evans_decline_1993], and popular opinions about the welfare state and redistribution [@Svallfors2006]. Regarding the latter, despite the claims of a displacement of the distributive struggle between classes by other group identities [@clark_declining_1993;@inglehart_culture_1990], past decade research has persistently demonstrated that class positions indeed still explain economic attitudes towards the welfare state and redistribution [@lindh_class_2020, p. 421]. Thus, the claimed blurring of the class cleavage has not been strongly supported while the class differences have been evidenced in perceived class conflicts [@perez_class_2023;@edlund_democratic_2015], attitudes on market-based inequality [@andersen_preferences_2018;@lindh_public_2015], and redistributive preferences [@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010].

However, the study of redistributive preferences has been mainly focused on the individual and household situation. In contrast, less attention has been paid to the sociability structures embedded in class relations regarding social tie formation. In this regard, how people form their friendship and partnership ties tends to be homogeneous because of the tendency towards homophily in forming interpersonal networks [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Hereby, we argue that the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes. In this regard, recent research has argued that the degree of segregation in the composition of social ties is a window for learning about others' lifestyles, economic conditions, and worldviews [@mijs_inequality_2018;@otero_power_2023]. However, there has been no systematic dialogue regarding the role of social ties on preferences for redistribution from a social network perspective. Therefore, this research aims to contribute to the ongoing literature on class-based network segregation on attitudes toward redistribution [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022].

While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention must be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003]. This research aims to contribute to the literature on preferences toward redistribution from a relational perspective, focusing on how the structural properties of networks can influence people's preferences toward economic redistribution.

Our research seeks to provide answers to two key questions. Firstly, at the micro-level, what is the role of class-based network segregation on redistributive preferences? Secondly, at the macro-level, does economic inequality play a role in the relationship between class-based network segregation and redistributive preferences? To achieve this, we have analyzed data from the International Social Survey Program (ISSP) 2017, which comprises a sample of 31,191 individuals across 31 countries globally. In this regard, the ISSP has a unique opportunity to date since it includes information on interpersonal networks through the position generator, the core questions for class position allocation, and attitudes toward economic inequality and redistribution.

@@

@@

header <- list("Q1"=1:2,"Q2"=3:4,"Q3"=5:6,"Q4"=7:8,"Q5"=9:10)
screenreg(digclas1_Q05,stars = c(0.001, 0.01, 0.05, 0.1),single.row = F,include.ci = FALSE,custom.header = header,
       caption = paste("(\\#tab:table1-groups)","Fixed effects linear regression models for Class-based network segregation and Redistributive Preferences by Income Inequality Quintiles"),
       custom.note = "Note: Models include sampling weights. Gender, age and marital status are included as controls. Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,
       custom.coef.map = ccoef,
       symbol = "+",scalebox = 0.55,
      groups = list("Social Class (Ref.= Service Class)" = 3:4, "Household Income (Ref.= Tertile I)"= 6:8,  
                    # "Partner's Social Class (Ref.= Service Class)" = 12:14, 
                    "Homogeneity x Social Class"=10:11),
      custom.gof.rows = list("Country FE"=rep("Yes",10)),
      # include.loglik = FALSE,include.aic = FALSE,
      # custom.gof.names = c(NA,NA,"Num. groups","Var: Country (Intercept)","Var: Residual"),
       float.pos = "h!",
      # file = "table1.html"
       use.packages = F)
```

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | The escalating economic inequality and the current sanitary crisis have threatened social solidarity among citizens. However, these consequences have been unevenly experienced across the social structure where the upper and intermediate classes, benefiting from privileged access to social resources, have shown resilience. Conversely, working-class families have faced deteriorating material conditions, leading to a heightened sense of marginalization and increasing the demand for welfare support. Particularly, social resources through network ties have been suggested to have a direct link with an individual's welfare through instrumental and expressive outcomes, such as providing information or help in moments of need. Consequently, it has been argued that being socialy excluded not only plays a role in terms of social resources, but higher levels of segregation can bolster opinions, suggesting that being isolated from other social classes can polarise attitudes mainly in the working and upper-middle classes. In this regard, based on the ISSP 2017 -  Social Networks (N=31,191),  multilevel estimations suggest that the influence of network homogeneity is conditional to social class, where the working and intermediate classes with high homogeneous networks hold higher redistributive preferences than the upper classes. Additionally, income inequality has been shown to decrease the polarizing effect of network segregation, mainly among the upper classes, when overall inequality is high. In sum, this study highlights the role of both social class and network segregation together to understand redistributive preferences, as well as their implications for policies and cohesion.
  | 
  | **Keywords**: network segregation, redistributive preferences, economic inequality
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

The study of redistributive preferences and social class has been mainly focused on the individual and household situation. However, less attention has been paid to the sociability structures embedded in class relations regarding social tie formation. Moreover, friendship and partnership ties tend to be homogeneous due to the homophilous tendency in interpersonal network formation [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Thus, the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes [@Svallfors2006]. In this respect, recent research has argued that the degree of segregation in the composition of social ties can influence opinions about economic inequality as they function as a window for learning about the lifestyles, economic conditions, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. However, a knowledge gap in the literature still needs to be understood regarding the network mechanisms in forming individual opinions about governmental actions that aspire to achieve egalitarian economic distribution of resources and opportunities in society [@mccall_americans_2009].

The study of redistributive preferences and social class has been mainly focused on the individual and household situation. However, less attention has been paid to the sociability structures embedded in class relations regarding social tie formation. Moreover, friendship and partnership ties tend to be homogeneous due to the homophilous tendency in interpersonal network formation [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Thus, the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes [@Svallfors2006]. In this respect, recent research has argued that the degree of segregation in the composition of social ties can influence opinions about economic inequality as they function as a window for learning about the lifestyles, economic conditions, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. However, a knowledge gap in the literature still needs to be understood regarding the network mechanisms in forming individual opinions about governmental actions that aspire to achieve egalitarian economic distribution of resources and opportunities in society [@mccall_americans_2009].

Recent studies have argued that understanding class relations requires focusing on network composition as it represents cross-class embeddedness as a property of the social structure [@blau_macrosociological_1977]. In this regard, it has been argued that the class position of network ties can influence opinions on redistributive policies, but also the degree of segregation from other social classes in more homogeneous social networks can strengthen the class divide in redistributive preferences  [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention must be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003].

Comparative evidence about the contextual factors that affect the relationship between class-based network segregation and redistributive preferences is scarce. Previous research on the class-attitude link has suggested that the distribution of economic resources is one of the main societal characteristics that can affect the link between class relations and redistributive demands, as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. These studies have indicated that the affluent are more sensitive to changes in income inequality and support redistribution to a greater extent when inequality rises, while the working class remains relatively stable. In this discussion, the sociological moral economy perspective has suggested that the upper classes are more concerned about the erosion of the opportunity structure as a matter of distributive justice [@sachweh_why_2019]. In contrast, the political economy approach has suggested that the negative consequences of inequality (e.g., criminality) can motivate greater support for redistribution among the upper classes [@dimick_altruistic_2017].

It has been previously established that wider economic gaps lead to segregated forms of social participation and status composition of social networks [@pichler_patterns_2007;@otero_space_2022]. Indeed, people with lower economic backgrounds have fewer chances to participate in social activities and form diverse social networks. Therefore, it has been suggested that material inequality reinforces the stratified access to social activities and the social distance between social classes, resulting in an increasing marginalization of people with lower resources and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023].

Only a few studies have examined the role of network structure from a comparative perspective, employing a well-established instrument as the position generator [@hadler_issp_2020;@sapin_issp_2020]. Despite previous evidence, to our knowledge, no research has stressed the role of economic inequality in the relationship between network segregation and redistributive preferences from a cross-national perspective.

Therefore, our study is guided by two key research questions. First, at the micro-level, we explore the role of class-based network segregation in shaping redistributive preferences. Secondly, at the macro level, we investigate whether economic inequality conditions the relationship between class-based network segregation and redistributive preferences. To address these questions, we have conducted an analysis of data from the International Social Survey Program (ISSP) 2017, which includes a sample of 32,529 individuals across 31 societies. The ISSP dataset is particularly valuable for our study as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

The class cleavage has not been strongly supported while the class differences have been evidenced in perceived class conflicts (Edlund & Lindh, 2015; Pérez, 2023), attitudes on market-based inequality (Andersen &Yaish, 2018; Lindh, 2015), and redistributive preferences (Brooks & Svallfors, 2010; Curtis & Andersen, 2015; Langsæther & Evans, 2020).

@@

header <- list("Q1"=1:2,"Q2"=3:4,"Q3"=5:6,"Q4"=7:8,"Q5"=9:10)
texreg(digclas1_Q05,stars = c(0.001, 0.01, 0.05, 0.1),single.row = F,include.ci = FALSE,custom.header = header,
       caption = paste("(\\#tab:table1-groups)","Fixed effects linear regression models for Class-based network segregation and Redistributive Preferences by Income Inequality Quintiles"),
       custom.note = "Note: Models include sampling weights. Gender, age and marital status are included as controls. Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,
       custom.coef.map = ccoef,
       symbol = "+",scalebox = 0.55,
      groups = list("Social Class (Ref.= Service Class)" = 3:4, "Household Income (Ref.= Tertile I)"= 6:8,  
                    # "Partner's Social Class (Ref.= Service Class)" = 12:14, 
                    "Homogeneity x Social Class"=10:11),
      custom.gof.rows = list("Country FE"=rep("Yes",10)),
      # include.loglik = FALSE,include.aic = FALSE,
      # custom.gof.names = c(NA,NA,"Num. groups","Var: Country (Intercept)","Var: Residual"),
       float.pos = "h!",
      # file = "table1.html"
       use.packages = F)
```
